The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-10-15 article.tex
Changes by @richelbilderbeek

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\richel{Have you already looked up for a target journal?} 
\gio{Honestly I have literally no idea how to select a good journal for this kind of article.}
\richel{I suggest you ask Rampal}

@@

%Current limits in current tools.
\item However, currently available priors can be not suitable to analyze some specific datasets. 
%With this work we aim to test whether or not the implementation of a new prior model is beneficial to study a specific kind of diversification process.

@@

\item The Multiple birth hypothesis can be useful to explain a phenomenon that has always puzzled evolutionary biologists: what are the drivers of the diversification processes for those phylogenies that show an impressive amount of speciation events in relatively short times? 
The (constant-rate) birth-death (BD) model embodies the common assumption that 
only a single speciation event can occur at any given time.
The multiple-birth-death (MBD) model 
\richel{I feel MBSD (Multiple Birth Single Death) may be a better name: extinctions are still one at a time}
\gio{I always had the same doubt. The problem is that a good name should be, at the same time, both descriptive and short. I'll think about that.}
relaxes this assumption, allowing events in which 
large-scale environmental changes lead to a great number of species 
in relatively short time intervals. Such a hypothesis may be a better fit to describe the burst in in systems like cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria (\cite{janzen2016}, \cite{janzen2017}).

@@

\gio{see the razzo pipeline file: I am not sure what is the best way to do so}.

\begin{comment}
We want trees from the two models to contain the same amount of information, 
i.e. the same (expected) number of DNA mutations and 
the same number of taxa at the present.
To obtain these twin trees, 
\end{comment}
\richel{
  I suggest to first start describing how to create the twin tree,
  calculation of the speciation rates, then the twin alignment.
}
\gio{@richel: both assumptions seem to make sense to me. 
  If possible I would run both and see which one performs better. For now
  let's stick to option 1 in the manuscript.
}
\richel{
  Sorry for the confusion: the previous comment was about the order
  of describing out methods in this manuscript. As of the methods,  
  I really meant to create twin trees with 
  both equal number of taxa and equal expected number of mutations.
}
\gio{The procedure is explained in the razzo pipeline markdown file. See there. CLOSED}

we impose
that the expected number of mutations in an MBD tree, $m_{MBD}$ equals
the expected number of mutations in a BD tree, $m_{BD}$:

@@

We first generate a set of MBD trees. For each of them we can measure the amount of mutations $m_{MBD}$.

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
\gio{
  I don't agree. It is expected if you use an expected number 
  of species over time. If you have the correct $n(t)$ in principle you 
  should be able to get the exact number of mutations.
}
\richel{
  If you have the correct $n(t)$ you can indeed calculate exactly what
  the number of mutations will be, if and only if the mutation rate is
  zero. Except for this trivial case, I am curious about how you
  calculate this. What is the precise number of mutations on a DNA
  sequence of 1000 base pairs over one time unit with an 
  expected average per-nucleotide
  mutation rate of once per time? I would say one expects on average 1000 mutations
  to occur. What would your claim be?
}
\gio{Now I agree. CLOSED}

with crown age $-T$ (with $T>0$) in fact is given by
\richel{
  above stood 'crown age -T'. 
  I feel that a crown age is a positive number,
  but I know you have had a good reason.
  Perhaps better would be to write something explicit like:
  t\_now - t\_crown = T.
  Looking forward for a better suggestion than mine :-)
}
\gio{
  We can specify that the crown age is $-T$ where $T$ is a positive number 
  (which should be already clear, I guess, 
  if we report it as $-T$). In my opinion this makes everything 
  nicer when we have to write down the integral which should, 
  in principle, be $\int_{-T}^{0} f(t-(-T))$ but using integral properties 
  can be rewritten as $\int_{0}^{T} f(t)$. Having a definite integral 
  starting from $0$ always looks better ^^.
}
\richel{
  I see, in a conversion you would say 'the crown age of the phylogeny is 
  minus ten million years', where I would the say 'the crown age of the 
  phylogeny is ten million years' to say exactly the same thing!
  To try to resolve this, I checked one (actually two) of Rampal's papers (
    'Diversity-dependence brings molecular phylogenies closer to agreement with the fossil record',
    by Rampal, Bart Haegeman, Tanja Stadler, Tracy Aze, Paul N. Pearson, Andy Purvis, Albert B. Phillimore
  ). Rampal avoids this point altogether: '... crown age $t_c = t_p − t_1$'
  and '... and crown ages (5, 10 and 15 Myr)'. I suggest we do the same in
  both writing and derivations: he is even more famous than us :-) 
}

@@

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}
\richel{
  I see you use angle bracket as a notation for the expected
  value. I usually see 'E(x)' as the expected value for 'x',
  and this is used at the beloved https://en.wikipedia.org/wiki/Expected\_value.
  What are the reasons you prefer the notation with the angle brackets?  
}
\gio{
  Because the capital E is inelegant, plebeian and boor. 
  And we want to be classy, don't we? :D
}
\richel{
  If the angle brackets are the new hip kids in mathematics town, sure,
  we'll use those!
}

@@

\richel{I suggest $n_{BD} = n_{MBD}$ and only 
change $\rho_{BD}$ to reach $<m_{MBD}> = <m_{BD}>$}
\gio{@Richel: Don't you think it might make more sense to set $\mu_{BD} = \mu_{MBD}$? What changes in the two model is the way we use to generate new species, not the way to remove them. Maybe one thing that's possible to do would be to make $\lambda_{BD}$ a function of time, a bit like Théo is doing in his comparison between DD and TD4 (which, by the way, seem to yield very different results). In case you are wondering the theory can be found in Caesar's master thesis.}
\richel{
  I fully agree to use the same extinction rates!
  The 'mu' used in the context of mutations (now 'rho') messed me up.
  I hope this is clear now. To recap: 
  (1) calculate the speciation rate of the twin tree as you wrote down excellently,
  (2) simulate a twin tree with same number of taxa,
  (3) calculate the mutation rates of the trees, 
    so their alignments contain as much information
}

\gio{My doubt is if we need to use $m_{MBD}$ for the single tree or the same
quantity averaged on the full MBD dataset $<m_{MBD}>$.
Do you think is better to use the
individual $m_{MBD}$ for each tree or the average across the whole
dataset?}
\richel{
  I think a per-tree calculation of the mutation rates 
  is the best we can do. As there was some noise between us above,
  I think the iteration after the next will allow us to get a better
  idea about this
}
\gio{
  I am still very debated about this. I actually would like to ask also Rampal's opinion.
}
\richel{
  Sounds like a good idea!
}

@@

\item So far we have simulated two datasets of trees under the two models: 
$\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{
  1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
}
\richel{
  I definitely get your point! 
  I would prefer something expressive over a number, perhaps a fancy letter, like
  for example $\mathbb{B}$ and $\mathbb{M}$ 
  for a Birth-Death and Multiple-Birth Death. 
}
\gio{
2) Writing down things to tidy up manuscript and thoughts I just realized that 
  actually the "model" is the same for the two posteriors. 
  In fact with the word "model" Maturana et al. refer to the prior 
  used (see eq. 1, $\pi(\theta | M)$). In such case are we sure we can actually 
  use the Bayes factor for model selection?
}
\richel{
  Correct! The assumed model of the posterior is a Birth-Death model. To do
  model selection, we could assume a different nucleotide substitution 
  model (GTR instead of JC69) and/or a different clock model (relaxed clock,
  instead of a strict clock). I hope this clears things up. I will add
  'generative model' for the model (BD or MBD) that created a phylogeny
  and 'inferential model' for the model used in generating a posterior.
}
}

\subsubsection{Option 1: nLTT statistics}
\item To compare the results for the two models we measure the inference 
error using the nLTT statistic between known/true tree and 
posterior/inferred trees (\cite{nltt}). 
To obtain such statistics the procedure is the following:

@@

\subsubsection{Option 2: Bayes Factor (BF)}

\richel{
  I removed the Bayes factor text. It is useful when letting BEAST2
  pick more/overly complex models and see if that more complex model fits the
  data better (penalized by its increased complexity, similar to the AIC). It has
  its uses, but I am unsure if we already want to discuss this now
  or first focus on the proper tree twinning
}
\gio{
  But after we describe the tree twinning we will have to describe 
  the method we use for model selection anyways. 
  BF and nLTT are mutually exclusive choices, right?
}
\richel{
  BF and nLTT are not mutually exclusive choices, but do complementary things:
  nLTT measures the difference between the 'true' tree and the inferred trees,
  which is the most important thing: we are interested in the error BEAST2
  makes when using an overly simple (single-)BD tree prior.
  With the BF we can compare BEAST2 posteriors that use (the same
  tree prior, but) different site models (e.g. JC69, the simplest, 
  versus GTR, the most complex). Although we know we generated the DNA
  alignment with JC69, there are some hints that the GTR model is the
  superior choice to do inference with.
}
\gio{
  The point is: what method do we choose? 
  According to the choice we have to describe that. 
  I'd also like to know Rampal's opinion on that. 
  In principle I like more the BF method cause it looks 
  more solid while the nLTT looks a bit like the last resort to me.
}
\richel{
  I hope this is now more clear. If not, I'll happily
  elaborate :-)
}

@@

\item
\richel{
  I guess you know I am a fan of the Open Science Framework,
  in which you first register you work before you do the experiment
  (note: I will do some small pilots to estimate the complete time
  of the experiment). I think it is the proper and superior science,
  which helps us against writing down bullshit stories after having
  obtained the results (e.g. 'We expected A and indeed found it!').
  It also helps me structure my work: first think
  deeply about the experiment, then do it (instead of the mixing
  up the two phases). What are your thoughts on that?
}
\gio{I agree on the principle but I don't fully agree on the open science procedure. Sometimes you do find something good while you're looking for something else. That happened many times in the history of science. This doesn't necessarily implies that we will end up p-hacking the results. I think we need to be open to new findings that might be not forecasted by our hypothesis. I think this is the good way to do science: be critical with our results but not imposing that reality will follow the theory.}
\richel{
  Looks like we are on the same page here: we both dislike making up bullshit 
  stories after the results are known. Great!
}
\item

>>>>

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\richel{Have you already looked up for a target journal?} 
\gio{Honestly I have literally no idea how to select a good journal for this kind of article.}
\richel{
  May I suggest we aim for Molecular Phylogenetics and Evolution, the same journal as the raket paper?
}

@@

%Current limits in current tools.
\item However, currently available priors can be not suitable to analyze some specific datasets. 
  With this work we aim to test whether or not the implementation of a new prior model 
  is beneficial to study a specific kind of diversification process.

@@

\item The Multiple birth hypothesis can be useful to explain a phenomenon that has always puzzled evolutionary biologists: what are the drivers of the diversification processes for those phylogenies that show an impressive amount of speciation events in relatively short times? 
The (constant-rate) birth-death (BD) model embodies the common assumption that 
only a single speciation event can occur at any given time.
The multiple-birth-death (MBD) model 
relaxes this assumption, allowing events in which 
large-scale environmental changes lead to a great number of species 
in relatively short time intervals. Such a hypothesis may be a better fit to describe the burst in in systems like cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria (\cite{janzen2016}, \cite{janzen2017}).

@@

\gio{see the razzo pipeline file: I am not sure what is the best way to do so}.
\richel{Looks fine to me!}

We want the MBD and twin BD trees to contain the same amount of information, 
i.e. the same number of DNA mutations and the same number of taxa at the present:

@@

The expected number of mutations $m$ of a phylogeny 
with crown age $-T$ (with $T>0$) in fact is given by
\richel{
  So one of use likes '-T', the other likes 'T'. How to resolve this?
}

@@

The parameter we'll tune is $\rho$ ... \richel{elaborate here :-)}

Since we cannot know $n_{BD}(t)$ before running simulations
we need to replace it with a proxy. 
For this reason we will use the average number of
species in time according to the BD model. 
It's well known that this is equal to \gio{insert proper citation}

@@

@@

\item So far we have simulated two datasets of trees under the two models: 
$\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{
  1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
}

\item To compare the results for the two models we measure the inference 
error using the nLTT statistic between known/true tree and 
posterior/inferred trees (\cite{nltt}). 
To obtain such statistics the procedure is the following:

@@

@@

\item

\item
